Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-19 icfp.tex
added Progress and partial derivations

We begin by presenting the operational semantics of $\lambda^\pi_\D$,
consisting of two lazy big-step semantics in the
style of Launchbury~\cite{10.1145/158511.158618}.

@@

The first semantics, presented in Figure~\ref{fig:opsem}, captures
lazy evaluation but ignores linearity information. The second
semantics is instrumented with type information and is thus
\emph{linearity aware}, resulting in a stuck state when a linear
variable is used more than once. We show that for well-typed terms,
the two semantics coincide.

@@

We follow an approach
similar to that of \cite{cite:linearhaskell}, defining a
linearity-aware operational semantics in which linear variables are
erased from 
the runtime environment, ensuring that any term violating linearity
will result in a stuck state. The semantics is also type-aware (for
the purposes of proving type safety), defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which
variables in $\Theta$ are used.

@@

\subsection{Results}

@@

\subsection{Type safety\label{sec:type-safety-meta}}

@@

@@

\begin{theorem}[Type Preservation]
  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) : \tau , \Sigma$.
  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$

@@

>>>>

We begin by presenting the operational semantics of $\lambda^\pi_\D$,
consisting of two lazy big-step semantics in the
style of Launchbury~\cite{10.1145/158511.158618}, one of which is
instrumented with linearity information, producing stuck states when
linearity is violated. We then show the
type safety  of $\lambda^\pi_\D$ via the standard progress  and type
preservation results (Section~\ref{sec:typesafe}), which in our setting entails that programs
preserve linearity. Finally, we show that multiple optimising
transformations such as those performed by GHC on (Linear) Core are
type preserving
(Section~\ref{sec:optimisations-preserve-types-meta}), and thus
linearity preserving.

@@

The first semantics, presented in Figure~\ref{fig:opsem}, captures
standard lazy evaluation but ignores linearity information. The second
semantics is significantly instrumented with type information and is thus
\emph{linearity aware}, resulting in a stuck state when a linear
variable is used more than once. The former is a more direct model of
an implementation of lazy evaluation, however it is not so convenient
to reason about formally. The latter semantics is augmented with the additional
information necessary to strengthen the induction hypothesis to make our
type safety proofs go through.
%
We show that for well-typed terms, the two semantics coincide.

@@

We follow an approach
similar to that of \cite{cite:linearhaskell}, defining a
linearity-aware operational semantics in which linear variables are
erased from 
the runtime environment, ensuring that any term violating linearity
will result in a stuck state. The semantics is also type-aware and
carries with it sufficient data to reconstruct typing derivations (for
the purposes of proving type safety). The semantics is defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which
bindings in $\Theta$ are used. This additional data is needed to
inductively show that the evaluation state is well-typed (and
linearity preserving).

@@

\subsection{Type Safety}\label{sec:typesafe}

@@

%\subsection{Type safety\label{sec:type-safety-meta}}

@@

 Following~\cite{Gunter1993APA,cite:linearhaskell}, we also consider
 partial derivations of our semantics in order to be able to state a
 progress result. Given our set of rules defining judgments of the
 form $a \Downarrow b$ with ordered premises, we define a total
 derivation of $a \Downarrow b$ as a tree in the standard way. A
 partial derivation $a \Downarrow?$ is either a root labelled with $a
 \Downarrow?$ or an application of a rule matching $a$ on the left of
 the evaluation arrow, where exactly one of the premises $a'
 \Downarrow?$ has a partial  derivation, all the premises to the left
 of $a' \Downarrow?$ have a total derivation, and the premises to the
 right of $a'\Downarrow?$ are not known yet.
 %

 By construction, in a partial derivation, there is only one premise
 $b\Downarrow?$ with no sub-derivation. We call $b$ the head of such a
 partial derivation.
We write $a \Downarrow^* b$ for the relation which holds when $b$ is
the head of some partial derivation with root $a \Downarrow?$.  We
write $a \Downarrow^* b$  the partial evaluation relation.

@@

\begin{theorem}[Type Preservation]
  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
  \tau , \Sigma$ or $\G ; \D \vdash (\Theta \mid e) \Downarrow^* (\Theta' \mid v) :
  \tau , \Sigma$.
  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$

@@

\begin{theorem}[Progress]
Let $\G ; \D \vdash \Theta \mid e : \tau,\Sigma$. Any partial derivation $\G; \D \vdash (\Theta \mid e)
\Downarrow?,\Sigma$ can be extended. 
\end{theorem}

\begin{proof}
 Essentially all cases are standard. The only potential issue is the
 (linear) variable rule.
It suffices to show that whenever a variable is evaluated it is
present in the evaluation environment. This situation is impossible
since the state
$\G ; \D \vdash \Theta \mid x : \tau , \Sigma$ where $x$ is absent
from $\Theta$ is not well-typed and so cannot be the head of a partial derivation.
\end{proof}
